### 2010-03-17 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: clean-up \note{}s. Update todo.
Desc@ICFP: apply \toDesc on tagged description.
Desc@ICFP: uniform Fig.~\ref
Desc@ICFP: uniformize bind and return style

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT [ \ldots ]$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

\note{Update to the factorised presentation}

@@

Of course, we must equip the resulting data-types with operations
delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
performs the r\^ole of ``return'', embedding variables into terms. The
``bind'' operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

@@

\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & \Mu{(\FreeMonad{D}{X})} \To
           (\V{X} \To \Mu{(\FreeMonad{D}{Y})}) \To
           \Mu{(\FreeMonad{D}{Y})} 
\end{array} \\
\subst\: D\: X\: Y\: x\: \sigma \mapsto
  \cata\: (\FreeMonad{D}{X})\: (\Mu{(\FreeMonad{D}{Y})})\: ???\: x
}\]

@@

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\FreeMonad{D}{X})}) \To
           \descop{\FreeMonad{D}{X}}{\Mu{(\FreeMonad{D}{Y})}} \To
           \Mu{(\FreeMonad{D}{Y})}                            \\
\end{array} \\
\begin{array}{@{}l@{\:\:\mapsto\:}l}
\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}     & \sigma\: x                   \\
\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
\end{array}
}\]

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives our
user the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the corresponding
monadic operation, ``bind'' and ``return''. Whereas the ``return'' is
a trivial variable former, the ``bind'' operation corresponds to the
substitution of variables by terms. We have shown how to implement it,
for any free monad.

@@

So, secondly, we have equipped this new data-type with its inherent
structure: the bind and return operations. We have been able to build
them as generic functions. Here, we rely on our ability to compute
over descriptions, thanks to the induction principle.

>>>>

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT [ \ldots ]$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

@@

Of course, we must equip the resulting data-types with operations
delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
performs the r\^ole of \return, embedding variables into terms. The
\bind operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

@@

\[\stk{
\begin{array}{@{}ll}
\subst : & \PITEL{\V{D}}{\TagDesc}
           \PI{\V{X}, \V{Y}}{\Set} \\
         & \Mu{(\toDesc{\FreeMonad{D}{X}})} \To
           (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{Y}})}) \To
           \Mu{(\toDesc{\FreeMonad{D}{Y}})} 
\end{array} \\
\subst\: D\: X\: Y\: x\: \sigma \mapsto
  \cata\: (\toDesc{\FreeMonad{D}{X}})\: 
          (\Mu{(\toDesc{\FreeMonad{D}{Y}})})\: 
          ???\: 
          x
}\]

@@

\[\stk{
\begin{array}{@{}ll}
\apply : & \PITEL{\V{D}}{\TagDesc} 
           \PI{\V{X}, \V{Y}}{\Set} \\
         & (\V{X} \To \Mu{(\toDesc{\FreeMonad{D}{X}})}) \To
           \descop{\toDesc{\FreeMonad{D}{X}}}{\Mu{(\toDesc{\FreeMonad{D}{Y}})}} \To
           \Mu{(\toDesc{\FreeMonad{D}{Y}})}
\\
\end{array} \\
\begin{array}{@{}l@{\:\:\mapsto\:}l}
\apply\: D\: X\: Y\: \sigma\: \pair{\DVar{}}{x}{}   & \sigma\: x                   \\
\apply\: D\: X\: Y\: \sigma\: \pair{\etag{t}}{ys}{} & \Con{\pair{\etag{t}}{ys}{}}
\end{array}
}\]

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives our
user the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the corresponding
monadic operation, \bind and \return. Whereas the \return is a trivial
variable former, the \bind operation corresponds to the substitution
of variables by terms. We have shown how to implement it, for any free
monad.

@@

So, secondly, we have equipped this new data-type with its inherent
structure: the \bind and \return operations. We have been able to
build them as generic functions. Here, we rely on our ability to
compute over descriptions, thanks to the induction principle.

### 2010-03-17 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix figure of tagged indexed description.
Desc@ICFP: uniformize bind and return style

\[\stk{
 \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}}
 \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
 \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})})  \\
}\]

@@

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic ``bind''. This is not surprising as we are defining a
first-order syntax with variables: our data-type enjoys more structure
than what we are given. In particular, this is reminiscent to a free
monad, where $\Eval{}$ is the ``return'' introducing
variables. Instead of implementing $\substH$ in this special case, we
shall implement it in a generic setting, and express our example in
that setting. However, before entering the realm of generic
structures, let us further explore the universe of indexed
descriptions.

@@

Just as in the universe of descriptions, this construction comes with
an obvious ``return'' and a substitution operation, the ``bind''. Its
type signature is the following:

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatypes $\HExprD$ and $\HExprVarD$ enjoy a monadic structure. We had
identified the variable substitution operation as the ``bind'' of a
free monad. The definition of $\substI$ above confirm our intuition:
using $\substI$, we should easily obtain $\substH$. To do so, we first
have to massage the definition of our data-type, to exhibit its
monadic structure.

As previously mentioned, we identify $\Eval{}$ as the ``return'' of
the free monad, while the other components are the action of the
monad. As a result, the definition is similar to $\HExprD$ presented
in Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
replaced by $\HExprAFreeD$:

@@

Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
nothing but the ``return''. The remaining constructors are simply the
signature functor, trivially indexed by $\Unit$. Hence, we describe
this signature functor by the following code:

>>>>

\[\stk{
 \TagIDesc{I}  \mapsto \TIMES{\ATagIDesc{I}}{\ITagIDesc{I}} \\
 \ATagIDesc{I} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{I} \spi{\V{E}}{(\LAM{\_} \IDesc{I})}) \\
 \ITagIDesc{I} \mapsto \SIGMA{\V{F}}{I \To \EnumU} (\PI{\V{i}}{I} \spi{(F\: i)}{(\LAM{\_} \IDesc{I})}) 
}\]

@@

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic \bind. This is not surprising as we are defining a first-order
syntax with variables: our data-type enjoys more structure than what
we are given. In particular, this is reminiscent to a free monad,
where $\Eval{}$ is the \return introducing variables. Instead of
implementing $\substH$ in this special case, we shall implement it in
a generic setting, and express our example in that setting. However,
before entering the realm of generic structures, let us further
explore the universe of indexed descriptions.

@@

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its type
signature is the following:

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatypes $\HExprD$ and $\HExprVarD$ enjoy a monadic structure. We had
identified the variable substitution operation as the \bind of a free
monad. The definition of $\substI$ above confirm our intuition: using
$\substI$, we should easily obtain $\substH$. To do so, we first have
to massage the definition of our data-type, to exhibit its monadic
structure.

As previously mentioned, we identify $\Eval{}$ as the \return of the
free monad, while the other components are the action of the monad. As
a result, the definition is similar to $\HExprD$ presented in
Figure~\ref{fig:hexpr-full}, at the exception of $\HExprAD$ that is
replaced by $\HExprAFreeD$:

@@

Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
nothing but the \return. The remaining constructors are simply the
signature functor, trivially indexed by $\Unit$. Hence, we describe
this signature functor by the following code:
